Efficient pedigree recording for fast population genetics simulation

### 2017-11-16 forwards_paper.tex
removed some of the tables requirements

%%%%%%%%%%%%%%
\subsection*{Data structures:}

Here we describe in more detail the data structures underlying a tree sequence in \msprime.
These derive from those described by \citet{kelleher2016efficient},
but have been generalised and modified to remove redundancy.

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
and correspond to the vertices in the trees of a tree sequence.
Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
which determines the height of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.
This information is stored in the columns of a \textbf{Node Table},
having columns \emph{time}, \emph{population}, and \emph{flags}.
The column ``flags'' records other information
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.

Tree sequences are constructed by specifying over which segments of genome
which nodes inherit from which other nodes.  This information is stored by
recording the endpoints of each distinctly inherited ancestral segment,
the parental node, and the child node who inherited that segment.
As each such record describes particular edge across a swatch of trees in the tree sequence,
we call these \emph{edges}
and store them in the columns of an \textbf{Edge Table},
having columns \emph{left}, \emph{right}, \emph{parent}, and \emph{child}.

To record information about genetic variants we need to also record each mutation
and which nodes have inherited that mutation.
The tree structure takes care of inheritance -- all we need to do
is to record the highest node in the tree at the mutated site
that inherited that mutation.
As more than one mtuation may occur at a given site,
we separate this information into two tables,
first, the \textbf{Site Table} records for each variant site its \emph{position} and \emph{ancestral state}.
Here \emph{position} is a (floating point) position along the chromosome,
and \emph{ancestral state} is the genotype of the root of the tree at that site.
As for nodes, \textbf{site IDs} are given implicitly
by the (zero-based) index of the rows.
Then, we record in a \textbf{Mutation Table} the \emph{site}, \emph{node}, and \emph{derived state}
for each mutation.
Here ``site'' is the ID of the site at wihch this mutation occurred,
``node'' is the ID of the highest node that has inherited this mutation,
and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
unless another mutation occurs.
% \plr{Omit migrations?}

\paragraph{Definition of valid tables}
Here are the formal requirements for a set of nodes and edges to make sense,
and to allow ``msprime``'s algorithms to work properly.

To disallow time travel and multiple inheritance:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
For algorithmic reasons, we also require:
\begin{enumerate} \setcounter{enumi}{2}
    \item The leftmost endpoint of each chromosome is 0.0.
    \item Node times must be strictly greater than zero.
    \item Edges must be sorted in nondecreasing time order.
\end{enumerate}
Note that since each node time is equal to the amount of time since the \emph{birth} of the
corresponding parent, time is measured in clock time, not in meioses.

A forwards-time simulation does \textbf{not} naturally emit genealogical information
satisfying all these requirements.
However, \msprime{} implements two algorithms that will take a set of tables
satisfying only 1--4 and produce tables satisfying all requirements.
These are \texttt{sort\_tables}, to enforce sortedness requirements (without renumbering nodes),
and \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.

>>>>

%%%%%%%%%%%%%%
\subsection*{The table format}

Given a set of node and edge tables as described above,
there are a small set of requirements that ensure the tables
describe a valid tree sequence.
(There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guarenteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.  
Checking this is more difficult, but is implemented in \msprime{}'s API.

For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tabled emitted by a forwards-time simulation.
These can be rectified by first sorting (using \texttt{sort\_tables})
and then using \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.

Note that since each node time is equal to the amount of time since the \emph{birth} of the
corresponding parent, time is measured in clock time, not in meioses.
